Inference Experiments on some Stochastic Differential Equations \\ (miscellaneous toy models)

### 2019-08-28 LaTeX_code/Reports/old_reports/SDES_UQ.tex
updated tripizoidal

\title{Prediction of Power Generation}
\subtitle{ Waleed Alhaddad}

@@

\begin{frame}[label=guide]\frametitle{ Reader's Guide }

List of changes in this iteration:
\begin{itemize}
\item Implemented Beta transitions
\item Implemented time dependent forecast
\item Numerically solve for the first two moments
\item Time dependent path variability parameter
\item Convergence results of all the above
\end{itemize}

Next steps:
\begin{itemize}
\item Use real data and forecast.
\item Try exponential decay in the mean reversion.
\end{itemize}

Note :
\begin{itemize}
\item Green slides: possible future extensions
\item Red slides: notes to be removed
\end{itemize}

\begin{frame}\frametitle{ Toy model with Synthetic Data }
We generate paths from an SDE  with given parameters using Monte Carlo simulation and then fit the model to check the retrievability of the parameters. We consider the following cases:
\begin{itemize}
\item[1:]  Constant parameter Wiener process: finite and infinite horizon.
\item[2:]  Controlled Wiener process within an interval.
\item[3:]  Controlled Wiener process centered around a given function. 
\end{itemize}
\end{frame}

@@

{
\setbeamercolor{background canvas}{bg=green!30}

\begin{frame}\frametitle{High Frequency  SDE -  Possible Extensions}

\begin{itemize}
\item Try to understand the limit of when we need to switch to the high frequency from the low frequency regime. That is how it depends on the speed of the  mean reversion in our model.
\item Get the martingale estimator function of the Gaussian transition density explicitly.
\item State the optimality and efficiency of the estimators explicitly.
\item Obtain the stated theoretical convergence rate.
\item Try to show theoretically why in case 1.2 , the diffusion estimator is not consistent.
\item Understand and state the conditions under which we have a unique parameter estimation to our SDE.
\end{itemize}

\end{frame}
}

@@

\begin{frame}\frametitle{ Case 4: Following deterministic functions }
As we construct processes that follow or revert to deterministic function it is natural to state the SDE in terms of deviations from the deterministic function. Then, we have for $V_t= X_t - p_t$

\begin{equation}
\begin{split}
dV_t &= a(V_t; \bm{\theta}_t) dt + b (V_t; \bm{\theta}_t ) dW_t \quad t > 0 \\
V_0 & = v_0
\end{split}\label{main_V}
\end{equation}

where we choose $ a(V_t; \bm{\theta}_t) = - \theta_t V_t$ to have mean reversion with exponential speed in time controlled by some strictly positive parameter $\theta_t$. To keep the process within a unit box, we choose $ b (V_t; \bm{\theta} )= \sqrt{ 2 \alpha_t \theta_t p_t (1-p_t)X_t (1-X_t)   }$

\begin{frame}\frametitle{ Case 4: Implementation }
In the previous sections we relied on the Euler discretization of SDEs in order to obtain our estimators through the Maximum Likelihood approach. However, this introduces errors in the likelihood similar to the observed error in the wiener process following the sine wave in (\ref{fig:sine}). To prevent such artifacts, we solve for the moments of the process defined by the SDE in (\ref{main_V}). We have,

\begin{equation}
\begin{split}
\E [V_t^k] &=  \E \Big[ \Big( \int a(V_t; \bm{\theta}_t) dt + \int b (V_t; \bm{\theta}_t ) dW_t \Big)^k \Big] \quad t > 0 \\
V_0 & = v_0
\end{split}\label{main_V_expectations}
\end{equation}
For a Gaussian approximation, we are interested in the first two moments.

\begin{frame}\frametitle{ Case 4: Implementation - first moment }
The first moment $(k=1)$ is given by,

\begin{equation}
\begin{split}
 m_1(t) =\E V_t &=  \E \Big[  \int a(V_t; \bm{\theta}_t) dt \Big] + \E \Big[ \int b (V_t; \bm{\theta}_t ) dW_t  \Big] \\
 & = \E \Big[  \int a(V_t; \bm{\theta}_t) dt \Big]
\end{split}\label{main_V_1_moment}
\end{equation}
In our case, the drift is linear and is given by $ a(V_t; \bm{\theta}_t) = - \theta_t V_t$. then, we have 
\begin{equation}
\begin{split}
 & m_1(t) =\E V_t =     \Big[  \int - \theta_t \E V_t dt \Big] 
 \iff \frac{dm_1(t)}{dt}=    - \theta_t m_1(t) \ \ ; \quad m_1(0)=\E[V_0]= v_0
\end{split}\label{main_V_1_moment}
\end{equation}
This is an ODE which has solutions of the form, 
\begin{equation}
m_1(t) = v_0 \exp ( - \int_0^t \theta_s \ ds ) 
\end{equation}

\begin{frame}\frametitle{ Case 4: Implementation - second moment }
The second moment $(k=2)$ is given by,

\begin{equation}
\begin{split}
  =\E V_t^2 &=  \E \Big[ \Big( \int a(V_t; \bm{\theta}_t) dt + \int b (V_t; \bm{\theta}_t ) dW_t \Big)^2 \Big] \\
\end{split}\label{main_V_2_moment}
\end{equation}
Let $Y_t=g(V_t)=V_t^2$ then $g_t= 0 , \ g^{'} (V_t) = 2V_t$ and $g^{''} (V_t) = 2$. Using $It\hat{o}$ and Taylor, 
\begin{equation}
\begin{split}
dY_t =  \frac{\partial g}{\partial t} dt &+  \frac{\partial g}{\partial v} \Big(a (V_t; \bm{\theta}_t ) \ dt + b (V_t; \bm{\theta}_t ) dW_t  \Big)\\
& + \frac{1}{2} \frac{\partial^2 g}{\partial v^2} \Big(a (V_t; \bm{\theta}_t ) \ dt + b (V_t; \bm{\theta}_t ) dW_t  \Big)^2 + \mathcal{O}(dV_t^3)
\end{split}
\end{equation}
As $dt \to 0$, we have
\begin{equation}
\begin{split}
dY_t &=  \frac{\partial g}{\partial t} + \Big( a (V_t; \bm{\theta}_t ) \frac{\partial g}{\partial v}  + \frac{b^2 (V_t; \bm{\theta}_t )}{2} \frac{\partial^2 g}{\partial v^2} \Big) dt + b (V_t; \bm{\theta}_t ) \frac{\partial g}{\partial v} dW_t  \\
&= 2\Big( a (V_t; \bm{\theta}_t ) V_t  + \frac{b^2 (V_t; \bm{\theta}_t )}{2}  \Big) dt +  2V_t b (V_t; \bm{\theta}_t ) dW_t
\end{split}
\end{equation}

\begin{frame}\frametitle{ Case 4: Implementation - second moment }
The second moment $(k=2)$ is given by,

\begin{equation}
\begin{split}
\E [V_t^2] =  \E Y_t &= 2  \E\Big[ \int \Big( a (V_t; \bm{\theta}_t ) V_t  + \frac{b^2 (V_t; \bm{\theta}_t )}{2}  \Big) dt \Big] +  2 \E \Big[ \int V_t b (V_t; \bm{\theta}_t ) dW_t \Big]\\
&= 2  \E\Big[ \int \Big( a (V_t; \bm{\theta}_t ) V_t  + \frac{b^2 (V_t; \bm{\theta}_t )}{2}  \Big) dt \Big]\\
\end{split}
\end{equation}
In our case, the drift is $ a(V_t; \bm{\theta}_t) = - \theta_t V_t$ and the diffusion is given by $ b (V_t; \bm{\theta} )= \sqrt{ 2 \alpha_t \theta_t p_t (1-p_t)X_t (1-X_t)   } =  \sqrt{ 2 \alpha_t \theta_t p_t (1-p_t)(V_t + p_t) (1-V_t-p_t)   }$. Then,
\begin{equation*}
\begin{split}
m_{2}&(t)=\E [V_t^2] = 2  \int \Big( - \theta_t \E[V_t^2]  +  \alpha_t \theta_t p_t (1-p_t)\E[(V_t + p_t) (1-V_t-p_t) ] \Big) dt \\
&=2  \int \Big( - \theta_t \E[V_t^2]  +  \alpha_t \theta_t p_t (1-p_t)( p_t(1 - p_t) +(1-2p_t)\E[V_t] - \E[V_t^2] ) \Big) dt \\
&=2  \int \Big( -\E[V_t^2] [\theta_t + \alpha_t \theta_t p_t(1-p_t) ] + \E[V_t][\alpha_t \theta_t p_t (1-p_t) (1-2p_t)] + \alpha_t \theta_t p_t^2(1-p_t)^2 \Big) dt \\
\end{split}
\end{equation*}

\begin{frame}\frametitle{ Case 4: Implementation - second moment }
The second moment $m_2(t)$ is given by solving the ODE,

\begin{equation*}
\begin{split}
\frac{d m_{2}(t)}{dt}&=2 \Big( -m_{2}(t) [\theta_t + \alpha_t \theta_t p_t(1-p_t) ] + m_{1}(t)[\alpha_t \theta_t p_t (1-p_t) (1-2p_t)] + \alpha_t \theta_t p_t^2(1-p_t)^2 \Big) \\
\end{split}
\end{equation*}
where $m_2(0) = \E[V_0^2] = v_0^2$ and has a solution of the form,
\begin{equation*}
\begin{split}
&m_{2}(t) = m_{2}(0) \exp \Big( -2 \int_0^t  [\theta_\tau + \alpha_\tau \theta_\tau p_\tau(1-p_\tau) ] \ d\tau  \Big) + \\
& 2 \int\limits_{0}^{t}   \exp \Big( -2 \int_0^s  [\theta_\tau + \alpha_\tau \theta_\tau p_\tau(1-p_\tau) ] \ d\tau \Big) \Big( m_{1}(t-s)[\alpha_{t-s} \theta_{t-s} p_{t-s} (1-p_{t-s}) (1-2p_{t-s})]  \\
& + \alpha_{t-s} \theta_{t-s} p_{t-s}^2(1-p_{t-s})^2 \Big)  \  ds
\end{split}
\end{equation*}

\begin{frame}\frametitle{Following deterministic functions - Cases }

We consider the following cases:
\begin{itemize}
\item[4.1] Constant parameters $\bm{\theta}_t = \bm{\theta} $
	\begin{itemize}
	\item[4.1.1] reverting to a constant function
	\item[4.1.2] reverting to a deterministic function
	\end{itemize}
\item[4.2] Time dependent parameters $\bm{\theta}_t$
\end{itemize}

\begin{frame}\frametitle{ Case 4.1: Following deterministic functions - constant parameters }
In this case, the first moment  $m_1(t)$ reduces to,

\begin{equation}
m_1(t) = v_0 \exp ( -t\theta ) 
\end{equation}
And the second moment $m_2(t)$ reduces to,

\begin{equation*}
\begin{split}
&m_{2}(t) = m_{2}(0) \exp \Big( -2  [t \theta + \alpha \theta \int_0^t    p_\tau (1-p_\tau)  \ d\tau ] \Big) + \\
& 2 \int\limits_{0}^{t}    \exp \Big( -2 [s \theta + \alpha \theta \int_0^s    p_\tau (1-p_\tau)  \ d\tau ] \Big) \Big( m_{1}(t-s)[\alpha \theta p_{t-s} (1-p_{t-s}) (1-2p_{t-s})]  \\
& + \alpha \theta p_{t-s}^2(1-p_{t-s})^2  \Big) \  ds
\end{split}
\end{equation*}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }
For constant forecast $p_t = \frac{1}{2}$ the second moment reduce further,

\begin{equation*}
\begin{split}
&m_{2}(t) = m_{2}(0) \exp \Big( -2 t \theta (1 + \frac{\alpha }{4} ) \Big) +   \frac{\alpha \theta}{8} \int\limits_{0}^{t}  \exp \Big( -2 s \theta (1 + \frac{\alpha }{4} )\Big) \  ds \\
& =m_{2}(0) \exp \Big( -2 t \theta (1 + \frac{\alpha }{4} ) \Big) + \frac{\alpha }{16 (1+ \frac{\alpha}{4})} \Big(1- \exp ( -2 t \theta (1 + \frac{\alpha }{4} ) \Big) 
\end{split}
\end{equation*}
Then the variance is given by,
\begin{equation}
\V[V_t] =m_{2}(0) \exp \Big( -2 t \theta (1 + \frac{\alpha }{4} ) \Big) + \frac{\alpha }{16 (1+ \frac{\alpha}{4})} \Big(1- \exp ( -2 t \theta (1 + \frac{\alpha }{4} ) \Big)   - v_0^2 \exp ( -2t\theta ) 
\end{equation}
Note that the second moment $m_2(0)=v_0^2$ as the previous point $V_{t_{i-1}}=v_0$ deterministic.
%% CHANGE ALL zero lower bounds to t_{i-1} 
\end{frame}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }

To conclude, we have the following mean and variance functions,

\begin{equation}
\begin{split}
\E[V_{t_i}| V_{t_{i-1}}=v_0] &= v_0 \exp ( -(t_i - t_{i-1})\theta ) \\
 \V[V_{t_i}| V_{t_{i-1}}=v_0] &= \frac{\alpha }{16 (1+ \frac{\alpha}{4})} \Big(1- \exp ( -2 (t_i - t_{i-1}) \theta (1 + \frac{\alpha }{4} ) \Big) \\
\end{split}
\end{equation}

Next, we discretize  and obtain the likelihood.
\end{frame}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }
To obtain the likelihood, consider a set of N observations, $X= \{ X_{t_0^N} , X_{t_1^N} ,\ldots , X_{t_N^N} \}$ on intervals of $\Delta_N$. We have $V=X-\frac{1}{2}$ .  Then,

\begin{equation}
\begin{split}
\E[V_{i}| V_{i-1}=v_0] &= v_0 \exp ( -\Delta_N \theta )  \\
 \V[V_{i}| V_{i-1}=v_0] &= \frac{\alpha }{16 (1+ \frac{\alpha}{4})} \Big(1- \exp ( -2 \Delta_N\theta (1 + \frac{\alpha }{4} ) \Big)\\
\end{split}
\end{equation}

The Likelihood  of the transition density is given by,

\begin{equation}
\mathcal{L}(\theta;X) = \prod\limits_{j=1}^M\prod\limits_{i=1}^N\rho( {V_{j,i+1}|V_{j,i}}, \bm{\theta}) 
\end{equation}
And the Gaussian transition density,
\begin{equation*}
\mathcal{L}(\bm{\theta}; V) = \prod\limits_{j=1}^M \prod\limits_{i=1}^N  \frac{1}{\sqrt{2 \pi \V[V_{j,t_i}| V_{j,t_{i-1}}=v_0] }  } \exp\Big(\frac{-(V_{j,i+1}   - \E[V_{j,t_i}| V_{j,t_{i-1}}=v_0] )^2}{2\V[V_{t_i}| V_{j,t_{i-1}}=v_0]}\Big) 
\end{equation*}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }
We obtain the following convergence using the analytical moments obtained from the $V_t$ process and enforcing that the parameters $\theta$ and $\alpha$ are non-negative.

\begin{figure}
   \includegraphics[scale=0.22]{Figures/conv_M_moments_constant_func_sym.jpg}
    \includegraphics[scale=0.22]{Figures/conv_N_moments_constant_fun_symjpg.jpg}
  \caption{This is done for true parameters $\theta = 0.5$ and $\alpha=1$ using  L-BFGS-B optimization algorithm }
  %\label{fig:boat1}
\end{figure}
\end{frame}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }
Further, we note that the  inference on the drift needs special care in situations of high speed reversion to the mean. Such situations seem to occur in the following cases:
\begin{itemize}
\item  $ \theta > > 1$  and big enough $\alpha$
\item  $\Delta_N > > 1$

\end{itemize}
In both cases, we see a flattening in the likelihood in the direction of the drift parameter $\theta$ which  causes the optimizer to return inconsistent estimates of the drift parameter $\theta$. These two cases are applicable if the function or the process oscillates at a high frequency with respect to the other.

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }

\begin{figure}
    \includegraphics[scale=0.22]{Figures/sample_moments_constant_func_1_01.pdf}
    \includegraphics[scale=0.22]{Figures/sample_moments_constant_func_10_01.pdf}
      \includegraphics[scale=0.22]{Figures/sample_moments_constant_func_100_01.pdf}

    \includegraphics[scale=0.22]{Figures/conv_M_moments_constant_func_1_01.pdf}
    \includegraphics[scale=0.22]{Figures/conv_M_moments_constant_func_10_01.pdf}
      \includegraphics[scale=0.22]{Figures/conv_M_moments_constant_func_100_01.pdf}

          \includegraphics[scale=0.22]{Figures/conv_N_moments_constant_fun_1_01.pdf}
    \includegraphics[scale=0.22]{Figures/conv_N_moments_constant_fun_10_01.pdf}
      \includegraphics[scale=0.22]{Figures/conv_N_moments_constant_fun_100_01.pdf}

  \caption{We can see how an increase of $\theta >>1$ with big enough $\alpha$ causes sharp oscillations that make the inference of  $\theta$ unfeasible.  }
  %\label{fig:boat1}
\end{figure}
\end{frame}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }
In the second case of increasing $\Delta_N$, we see the following flattening of the likelihood in the direction of $\theta$.
\begin{figure}
    \includegraphics[scale=0.5]{Figures/dt_increasing_moments.pdf}

  \caption{We can see how an increase of time intervals $\Delta_N >>1$  causes a flattening of the likelihood making the valley of the optimal $\theta$ shallower  until the  inference of  $\theta$ becomes unfeasible.  }
  %\label{fig:boat1}
\end{figure}
\end{frame}

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }

Question: How can we quantify the mean reversion of an SDE   ?  \\

The reversion to the mean is not only determined by the parameter $\theta$ but also by the diffusion term. We can see this in the following example,

\begin{equation}
dX_t = (1+X_t)dt + (1+X_t^2) dW_t
\end{equation}

which at first glance does not seem mean reverting, but in fact,  it is mean reverting. The reversion is driven  solely  by diffusion term in this case.  This is a case of volatility induced stationary, as the non-linearity in the diffusion term are responsible for the mean reversion. A similar non-linearity is present in our model and quantifying the effective mean reversion  will help determine if we are in the convergent regime  where our estimator of the drift is consistent or not.

Regarding, the specific application to wind forecasts this should not be an issue as the forecast and the recorded data are unlikely to oscillate around each other at a "high" frequency.

\begin{frame}\frametitle{ Case 4.1.1: Following a constant function - constant parameters }

Answer: Yes, it is true that such SDE's are stationary even though they are not "mean reverting". The  "mean reversion" or the "pull" is induced solely by the diffusion term. We can show that

\begin{equation}
dX_t = (1+X_t)dt + (1+X_t^2) dW_t
\end{equation} 
is stationary. Even though its associated process with constant diffusion $\sigma$
\begin{equation}
dX_t = (1+X_t)dt + \sigma dW_t
\end{equation}
is clearly non-stationary. We can do that using the tools we will introduce next to check and find the stationary distributions of our SDEs. Namely, using the scale and speed densities. 
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }
Consider an SDE,
\begin{equation}
\begin{split}
dX_t &= a(X_t; \bm{\theta}) dt + b (X_t; \bm{\theta} ) dW_t \quad t > 0 \\
X_0 & = x_0
\end{split}\label{main}
\end{equation}
We define the scale density $s(x)$,
\begin{equation}
s(x) = exp \Big[ \int^x - \frac{2  a(y; \bm{\theta})}{b^2 (y; \bm{\theta} )}\ dy \Big]
\end{equation}
And the speed density $m(x)$,
\begin{equation}
m(x) = \frac{1}{b^2 (y; \bm{\theta}) \  s(x)}
\end{equation}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }

A unique stationary distribution of the process $X_t$ exists if and only if 
\begin{equation}
\int_{x_0}^r s(x) \ dx = \infty \quad \int_{l}^{x_0} s(x) \ dx = \infty \quad \int_l^r m(x) \ dx < \infty
\end{equation}
where $x_0 \in (l,r)$ and $(l,r)$ is an interval where $b^2 (y; \bm{\theta} ) > 0$

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }
Considering the  SDE of a process following a constant forecast $p$,
\begin{equation}
\begin{split}
dX_t &= -\theta(X_t - p) dt + \sqrt{2 \theta \alpha p(1-p) X_t (1-X_t)} dW_t \quad t > 0 \\
X_0 & = x_0
\end{split}\label{main}
\end{equation}
We obtain the scale density $s(x)$ with integrability condition $\frac{p}{a} \leq -1 $ and $\frac{1-p}{a} \leq -1$
\begin{equation}
s(x) = exp \Big[ \int^x - \frac{2  a(y; \bm{\theta})}{b^2 (y; \bm{\theta} )}\ dy \Big] = exp \Big[ \int^x_{x_0}  \frac{y-p}{a y (y-1)}\ dy \Big]  \propto (1-x)^{\frac{1-p}{a}} x^{p/a}
\end{equation}
where $a= - \alpha p(1-p) <0$.Then the speed density $m(x)$ becomes, 
\begin{equation}
m(x) = \frac{1}{b^2 (y; \bm{\theta}) \  s(x)} \propto (1-x)^{-\frac{1-p}{a}-1} x^{-\frac{p}{a}-1}
\end{equation}
We note that $(l,r) = (-\infty, \infty)$ is where $b^2 (y; \bm{\theta} ) >0$. Let $x_0 \in (l,r)$. Taking $x_0=1$ we see that
\begin{equation}
\int_{x_0}^r s(x) \ dx = \infty \quad \int_{l}^{x_0} s(x) \ dx = \infty \quad \int_l^r m(x) \ dx < \infty
\end{equation}
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }
Under the condition,
\begin{equation}
\begin{cases}
\frac{p}{a} \leq -1\\
\frac{1-p}{a} \leq -1\\
\end{cases}\iff
\min (p, 1-p) \geq -a \iff  \min (p, 1-p) \geq \alpha p(1-p)
\end{equation}
For example, if forecast is constant $ p = \frac{1}{2}$, the condition becomes $\alpha \leq 2$.

we conclude that there exists a unique stationary distribution of  $X_t$ and its proportional to the speed density,
\begin{equation}
m(x)\propto (1-x)^{-\frac{1-p}{a}-1} x^{-\frac{p}{a}-1}
\end{equation}
 which is the kernal of a Beta distribution with shape parameters
\begin{equation}
  Beta \ \Big( \frac{p}{\alpha (1-p)}, \frac{1}{\alpha p } \Big)
\end{equation}
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }
Now we consider the centered process $V_t= X_t - p$,
\begin{equation}
\begin{split}
dX_t &= -\theta V_t dt + \sqrt{2 \theta \alpha p(1-p) (V_t +p ) (1-V_t-p)} dW_t \quad t > 0 \\
X_0 & = x_0
\end{split}\label{main}
\end{equation}
We obtain the scale density $s(x)$,
\begin{equation}
s(x) = exp \Big[ \int^x - \frac{2  a(y; \bm{\theta})}{b^2 (y; \bm{\theta} )}\ dy \Big] = exp \Big[ \int^x_{x_0} \frac{y-p}{a y^2 + by +c}\ dy \Big]  
\end{equation}
where,
\begin{equation}
\begin{cases}
a&=  \alpha p(1-p)\\
b&=  \alpha p(1-p)(2p-1)\\
c&=  \alpha p^2(1-p)^2\\
\end{cases}
\end{equation}
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }
To simplfy, we choose the constant forecast to be $p=\frac{1}{2}$. Then we have,

\begin{equation}
\begin{cases}
a&=  \frac{\alpha}{4} \\
b&=  0 \\
c&=  \frac{\alpha}{16} \\
\end{cases}
\end{equation}
and the scale density becomes, 
\begin{equation}
\begin{split}
s(x) &= exp \Big[ \int^x - \frac{2  a(y; \bm{\theta})}{b^2 (y; \bm{\theta} )}\ dy \Big] = exp \Big[ \int^x_{x_0} \frac{y}{a y^2 +c}\ dy \Big] =exp \Big[ \int^x_{x_0} \frac{y}{\frac{\alpha }{16}+\frac{\alpha  y^2}{4}}\ dy \Big]\\
& \propto \left(4 v^2+1\right)^{2/\alpha } \quad \quad (\alpha >0)
\end{split}
\end{equation}
Then we have the speed density $m(x)$,
\begin{equation}
m(x) = \frac{1}{b^2 (y; \bm{\theta}) \  s(x)} \propto \left(4 v^2+1\right)^{-\frac{\alpha +2}{\alpha }}
\end{equation}
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }
We proceed to check if it is stationary.

We note that $(l,r) = (-\infty, \infty)$ is where $b^2 (y; \bm{\theta} ) >0$. Let $x_0 \in (l,r)$. Taking $x_0=0$, we see that
\begin{equation}
\int_{x_0}^r s(x) \ dx = \infty \quad \int_{l}^{x_0} s(x) \ dx = \infty \quad \int_l^r m(x) \ dx < \infty
\end{equation}
So we conclude that it is a stationary distribution of $V_t$ and its unique.
Hence, the kernal of the invarient distribution of the process $V_t$ with constant forecast $p=\frac{1}{2}$ is $\pi(v)= \frac{\sqrt{\pi} \  \Gamma (\frac{1}{2}+ \frac{2}{\alpha})}{\Gamma(\frac{2+\alpha}{\alpha})}\left(1+4 v^2\right)^{-\frac{\alpha +2}{\alpha }}$. Note that it almost resembles a t-student distribution.
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Theoretical Background }

\begin{figure}
    \includegraphics[scale=0.9]{Figures/invarient_Vt.pdf}
  \caption{ The stationary distribution $\pi(v)$ of the process $V_t$ is symmetric with heavy tails.}
  %\label{fig:boat1}
\end{figure}

\begin{frame}\frametitle{Following deterministic functions - Cases }

In an improved model, we incorperate  Beta density transitions. We repeat the same cases as follows:
\begin{itemize}
\item[5.1] Constant parameters $\bm{\theta}_t = \bm{\theta} $
	\begin{itemize}
	\item[5.1.1] reverting to a constant function
	\item[5.1.2] reverting to a deterministic function
	\end{itemize}
\item[5.2] Time dependent parameters $\bm{\theta}_t$
\end{itemize}

However, first lets discuss why and how do we incorperate the Beta transition densities.

\begin{frame}\frametitle{ Case 5.1.1: Following a constant function - constant parameters }
Recall that both $X_t$ and $V_t$ are controlled processes that are constrained within a unit box. The symmetry of the Gaussian transition density of these two processes indicate that we have some probability outside the unit box once the forecast touches the boundry. There are many ways to fix this behaviour:
\begin{enumerate}
\item Use a Beta transition density for all the transitions while matching the moments of the Gaussian transition density.
\item Use a Beta transition density only when the transition density of the process touches the boundry of the unit box.This depends on the width and position of the transition density.
\begin{enumerate}
\item match the parameters of the Beta transition with the moments of the Gaussian transition.
\item Not to match the moments of the Gaussian and introdude the beta parameters as two additional hyperparameters (increasing the degrees of freedom).
\end{enumerate}
\end{enumerate}

\begin{frame}\frametitle{ Case 5.1.1: Following a constant function - constant parameters }

If we like to enforce that the Beta transition matches the moments of the Gaussian, then
we can solve for the paramters of the Beta distribution in terms of the parameters of the Gaussian transition density.

First, lets consider the non-centered process $X_t$. In that case, we choose standard Beta distribution which has support on $x \in [0,1]$. Then solving for its parameters in terms of the parameters of the Gaussian transition denisty, to obtain

\begin{equation} 
\alpha = \Big(  \frac{1-\tilde{\mu}}{\tilde{\sigma}^2} -\frac{1}{\tilde{\mu}}   \Big)\tilde{\mu}^2 \quad \quad \beta =  \alpha \Big(\frac{1}{\tilde{\mu}} -1 \Big)
\end{equation}
where $\tilde{\sigma}^2$ is the variance of the Gaussian transition density of the process and $\tilde{\mu}$ is its mean.
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Following a constant function - constant parameters }
To ensure the existence of such a Beta distribution with the parameters required
\begin{equation} 
\alpha = \Big(  \frac{1-\tilde{\mu}}{\tilde{\sigma}^2} -\frac{1}{\tilde{\mu}}   \Big)\tilde{\mu}^2 \quad \quad \beta =  \alpha \Big(\frac{1}{\tilde{\mu}} -1 \Big)
\end{equation}
we have the followind conditions by the definition of the standard Beta distribution:
\begin{enumerate}
\item $x \in [0,1]$.
\item Its shape parameters $\alpha$ and $\beta$ must be positive definite.
\end{enumerate}
Solution (this only applies to Euler realizations due to time discretization):
\begin{enumerate}
\item Discard realizations $X_t$ outside the range $(0,1)$.
\item Ensure that $\alpha >0$ and $\beta >0 $ with high probability.
\end{enumerate}
The first condition is automatically satisfied with high probability. The second condition introduces the following restrictions:
\begin{equation}
\alpha >0 ,\ \beta >0 \iff  0< \tilde{\mu}_t < 1  \ \text{and} \  \tilde{\sigma}_t^2 < \tilde{\mu}_t - \tilde{\mu}_t^2
\end{equation}
while the first restriction $ 0< \tilde{\mu}_t < 1$ holds with very high probability. The second restricts the second moment of the process $X_t$ not to exceed its mean,  i.e $\E[X_t^2] \leq \tilde{\mu}_t \quad \forall t$
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Following a constant function - constant parameters }
Next, we consider the centered process $V_t$. Since the process $V_t$ is  inside the range $(-1,1)$ with high probability, the standard Beta distribution cannot model its transitions. Therefore, we translate the distribution to have support on $[a,b]=[-1,1]$.  We use the linear transformation $y= g(x)= a + (b-a) x $ then we have
\begin{equation}
\  g^{-1}(y)= \frac{a-y}{b-a} \quad \text{and} \quad \Big| \frac{\partial g^{-1}(y)}{\partial y }  \Big| = \frac{1}{b-a}
\end{equation}
And the general Beta distribution becomes, 
\begin{equation}
f_Y(y)= f_X(g^{-1}(y) ) \Big| \frac{\partial g^{-1}(y)}{\partial y }  \Big| = \frac{1}{B(\alpha, \beta) (b-a)} \Big(\frac{a-y}{b-a}\Big)^{\alpha -1}\Big(1-\frac{a-y}{b-a} \Big)^{\beta -1}
\end{equation}
and we have,
\begin{equation}
\E[Y]= a + (b-a) \E[X] = a + (b-a) \frac{\alpha}{\alpha + \beta}
\end{equation}
\begin{equation}
\V[Y]= (b-a)^2 \V[X] = (b-a)^2 \frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)}
\end{equation}
\end{frame}

\begin{frame}\frametitle{ Case 5.1.1: Following a constant function - constant parameters }
Next, we need to solve the system in terms of the mean and variance of the process $V_t$.
\begin{equation}
\begin{cases}
\tilde{\mu}_t=a + (b-a) \frac{\alpha}{\alpha + \beta}\\
\tilde{\sigma}_t^2= (b-a)^2 \frac{\alpha \beta}{(\alpha + \beta)^2 (\alpha + \beta + 1)}\\
\end{cases}
\end{equation}
to obtain, 
\begin{equation}
\alpha = \frac{(\tilde{\mu}_t-a) \left(\tilde{\mu}_t^2-\tilde{\mu}_t (a+b)+\tilde{\sigma}_t^2+a b\right)}{\tilde{\sigma}_t^2 (a-b)}
\end{equation}

\begin{equation}
\beta = \frac{(\tilde{\mu}_t-b) \left(\tilde{\mu}_t^2-\tilde{\mu}_t (a+b)+\tilde{\sigma}_t^2+a b\right)}{\tilde{\sigma}_t^2 (b-a)}
\end{equation}
then for our interval of interest $[a,b]=[-1,1]$, we have

\begin{equation}
\alpha = - \frac{(1+\tilde{\mu}_t )(\tilde{\mu}_t^2 + \tilde{\sigma}_t^2 -1)}{2 \tilde{\sigma}_t^2} \quad \quad \beta =  \frac{(\tilde{\mu}_t-1 )(\tilde{\mu}_t^2 + \tilde{\sigma}_t^2 -1)}{2 \tilde{\sigma}_t^2} \label{param_transformed_beta}
\end{equation}

>>>>

\title{Inference  Experiments on some Stochastic Differential Equations \\ (miscellaneous toy models)}
\subtitle{ Waleed Alhaddad}

@@

%\begin{frame}[label=guide]\frametitle{ Reader's Guide }
%
%List of changes in this iteration:
%\begin{itemize}
%\item Matched data with every forecast
%\item Normalized according to installed power of each period
%\item Interpolated the forecast using cubic splines.
%
%\end{itemize}
%
%Next steps:
%\begin{itemize}
%\item upgrade the numerical integrator and use the interpolated forecast
%\item Check performace using the interpolated forecasts and data
%\end{itemize}
%
%Note :
%\begin{itemize}
%\item Green slides: possible future extensions
%\item Red slides: notes to be removed
%\end{itemize}
%
%\end{frame}

\begin{frame}\frametitle{ Toy models }
We generate paths from an SDE  with given parameters and then fit the model to check the retrievability of the parameters. We consider the following cases:
\begin{itemize}
\item[1:]  Constant parameter Wiener process: finite and infinite horizon.
\item[2:]  Controlled Wiener process within an interval.
\item[3:]  Controlled Wiener process centered around a given function. 
\end{itemize}
\end{frame}

@@

{
%\setbeamercolor{background canvas}{bg=green!30}

%\begin{frame}\frametitle{High Frequency  SDE -  Possible Extensions}
%
%\begin{itemize}
%\item Try to understand the limit of when we need to switch to the high frequency from the low frequency regime. That is how it depends on the speed of the  mean reversion in our model.
%\item Get the martingale estimator function of the Gaussian transition density explicitly.
%\item State the optimality and efficiency of the estimators explicitly.
%\item Obtain the stated theoretical convergence rate.
%\item Try to show theoretically why in case 1.2 , the diffusion estimator is not consistent.
%\item Understand and state the conditions under which we have a unique parameter estimation to our SDE.
%\end{itemize}
%
%\end{frame}
%}

@@

%\againframe{guide}
